A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-05-29 sections/Discussion.tex
further updates
edits in 5.3

\jimmy{TODO: update the summary}

@@

\subsection{Lexically-scoped type variables}
We further extend the type system to support a simple extension,
lexically-scoped type variables~\cite{scoped-type-variables}.
We have a complete Abella formalization for this extension proving
all the metatheory discussed in the paper.

From a
practical point of view, such extension enables the programmer to
refer to type variables in a generic function signature when
implementing that function.  For example,
$$(\lam{\lam[y] (x:a)}) : \all[a~b] a \to b \to a$$
has an inner annotation $(x:a)$ that refers to the outer generic type variable $a$.
This is not a surprising feature, since the declarative system already accepts these kind of programs
$$
\inferrule*[right=$\mathtt{DeclAnno}$]
{
    \Psi \vdash \all A \quad
    \inferrule*[right=$\mathtt{Decl\forall I}$]
    {\Psi,a \vdash e \Lto A}
    {\Psi\vdash e\Lto \all A}
}
    {\Psi\vdash (e: \all A)\To \all A}
$$

But there are issues about the well-formedness condition.
Normally $\Psi \vdash (e : A)$ is defined by $\Psi \vdash e$ and $\Psi \vdash A$.
However, when $A = \all A'$, the type variable $a$ is not introduced to the context of $e$,
therefore $\Psi \vdash e$ is not derivable.

To address the problem,
we add a syntax that explicitly bind the generic type variable into a function and its annotation.
The syntax of expression is extended with one more constructor:
$$
\begin{array}{l@{\qquad}lcl}
\text{Expressions}\qquad&e &::=&\quad x \mid () \mid \lam e \mid e_1~e_2 \mid (e:A) \mid \Lambda a.~e:A
\end{array}
$$

The new type-lambda syntax $\Lambda a.~e:A$ actually annotates the generic function $e$ with $\all A$,
while making $a$ visible inside the body of the function.
The well-formedness judgments are updated naturally:
$$
\inferrule*[right=$\mathtt{wf_d}\Lambda$]
    {\Psi,a \vdash e \quad \Psi,a \vdash A}
    {\Psi \vdash \Lambda a.~e:A}
\qquad
\inferrule*[right=$\mathtt{wf\_}\Lambda$]
    {\Gm,a \vdash e \quad \Gm,a \vdash A}
    {\Gm \vdash \Lambda a.~e:A}
$$

@@

In practice, programmers would not write the syntax $\Lambda a.~e : A$ directly.
The ScopedTypeVariables extension of Haskell is effective only
when the type signature is explicitly universally quantified
(the compiler translates into an expression similar to $\Lambda a.~e : A$);
otherwise the program means the normal syntax $e : \all A$,
and one may not later refer to the type variable $a$.

We also obtained all three desired properties for the extended system,
namely soundness, completeness and decidability.

>>>>

@@

\subsection{Lexically-scoped type variables}
We fhave urther extended the type system with support for
lexically-scoped type variables~\cite{scoped-type-variables}.
Our Abella formalization for this extension proves
all the metatheory we discuss here.

From a
practical point of view, this extension allows the implementation of a function
to refer to type variables from its type signature. For example,
$$(\lam{\lam[y] (x:a)}) : \all[a~b] a \to b \to a$$
has an annotation $(x:a)$ that refers to the type variable $a$ in the type signature.
This is not a surprising feature, since the declarative system already accepts similar programs
$$
\inferrule*[right=$\mathtt{DeclAnno}$]
{
    \Psi \vdash \all A \quad
    \inferrule*[right=$\mathtt{Decl\forall I}$]
    {\Psi,a \vdash e \Lto A}
    {\Psi\vdash e\Lto \all A}
}
    {\Psi\vdash (e: \all A)\To \all A}
$$

The main issue is the well-formedness condition.
Normally $\Psi \vdash (e : A)$ follows from $\Psi \vdash e$ and $\Psi \vdash A$.
However, when $A = \all A'$, the type variable $a$ is not in scope at $e$,
therefore $\Psi \vdash e$ is not derivable.

To address the problem, we add a new syntactic form that explicitly binds a
type variable simulatenously in a function and its annotation.
$$
\begin{array}{l@{\qquad}lcl}
\text{Expressions}\qquad&e &::=&\quad \ldots \mid \Lambda a.~e:A
\end{array}
$$

This new type-lambda syntax $\Lambda a.~e:A$ actually annotates its body $e$ with $\all A$,
while making $a$ visible inside the body of the function.
The well-formedness judgments are extended accordingly:
$$
\inferrule*[right=$\mathtt{wf_d}\Lambda$]
    {\Psi,a \vdash e \quad \Psi,a \vdash A}
    {\Psi \vdash \Lambda a.~e:A}
\qquad
\inferrule*[right=$\mathtt{wf\_}\Lambda$]
    {\Gm,a \vdash e \quad \Gm,a \vdash A}
    {\Gm \vdash \Lambda a.~e:A}
$$

@@

In practice, programmers would not write the syntax $\Lambda a.~e : A$ directly.
The ScopedTypeVariables extension of Haskell is effective only
when the type signature is explicitly universally quantified
(which the compiler translates into an expression similar to $\Lambda a.~e : A$);
otherwise the program means the normal syntax $e : \all A$
and may not later refer to the type variable $a$.

We have proven all three desired properties for the extended system,
namely soundness, completeness and decidability.
